Poisson approximate likelihood compared to the particle filter

### 2024-09-14 ms.Rnw
compile to pdf, sbh_data

<<benchmark,echo=FALSE>>=
arma_benchmark <- function(x,order=c(2,0,1)){
  arima(log(x + 1), order = order)$loglik - sum(log(x+1))
}

sum(apply(realdat,2, function(x)  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))))

sbh_data <- read.csv("sbh_data.txt",sep=" ",skip=3)[,-1]
sbh_benchmark <- sum(apply(sbh_data,2, function(x)
  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
))

wwr_benchmark <- sum(apply(realdat,2, function(x)
  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
))

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the pomp R package \citet{pomppackagepaper}. The first two lines are from WWR, Table~1, and lines 3--5 are our own computations. We used 50000 particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL. PF was maximized using iterated filtering, and PAL was maximized using coordinate gradient descent.
} 
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{lllr} 
  \hline
  & Method & Result & AIC  \\
  \hline
  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Data, model and parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Data, model and parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),tab_signif)}
  \\
  5. & PF & Changing initial values &
     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),tab_signif)}
  \\
  6. & PAL & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & Data from WWR &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}

>>>>

<<benchmark,echo=FALSE>>=
set.seed(123)

sbh_data <- read.csv("sbh_data.txt",sep=" ")[,-c(1)]
sbh_benchmark <- sum(apply(sbh_data,2, function(x)
  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
))

wwr_benchmark <- sum(apply(dat[,-1],2, function(x)
  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
))

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{AIC for the OvOv rotavirus model, computed using two filtering methods. PAL is the Poisson approximate likelihood, implemented using the code of WWR. PF is the particle filter, implemented using the pomp R package \citet{pomppackagepaper}. The first two lines are from WWR, Table~1, and lines 3--5 are our own computations. We used 50000 particle for both PF and PAL. PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL. PF was maximized using iterated filtering, and PAL was maximized using coordinate gradient descent.
} 
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{lllr} 
  \hline
  & Method & Result & AIC  \\
  \hline
  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
  \\
  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
  \\
  3. & PAL & Data, model and parameters from WWR &
    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
  \\
  4. & PF & Data, model and parameters matching 3. &
    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),tab_signif)}
  \\
  5. & PF & Changing initial values &
     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),tab_signif)}
  \\
  6. & PAL & Data, model and parameters matching 5.  &
     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
  \\   
  7. & ARMA & Sec.~7.3.2 of WWR & 23043
  \\
  8. & ARMA & Data from WWR &
    \Sexpr{myround(wwr_benchmark,tab_signif)} 
  \\
 \hline
\end{tabular}
\end{table}
